---
title: "Week 9 · Structure Analysis · Self-Similarity Matrices"
author: "John Ashley Burgoyne"
date: "27 February 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Week 9 · Structure Analysis · Self-Similarity Matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Set-up

```{r}
library(tidyverse)
library(spotifyr)
library(compmus)
```

In order for the code below to run, it is also necessary to set up Spotify login credentials for `spotifyr`.

## New functions

Last week, we used several custom functions to work with the Spotify API:

  - `get_tidy_audio_analysis` to load audio analyses from Spotify, one track at a time
  - `compmus_normalise` to normalise audio features using common techniques, including:
      - `manhattan`
      - `euclidean`
      - `chebyshev`
  - `compmus_long_distance` to compare to series of audio features against each other using common distance metrics, including:
      - `manhattan`
      - `aitchison`
      - `euclidean`
      - `cosine`
      - `angular`
  
This week, we have a few new custom functions. 

  - `compmus_align` aligns two levels of structure with each other, e.g., Spotify segments with beats or bars.
  - `compmus_summarise` helps to summarise features within higher levels of structure, including:
      - `mean`
      - `acentre` [Aitchison centre]
      - `rms` [root mean square]
      - `max`

## Common combinations

| Domain                      | Normalisation | Distance  | Summary Statistic |
| ----------------------------|---------------|-----------|-------------------|
| Non-negative (e.g., chroma) | Manhattan     | Manhattan | mean              |
|                             |               | Aitchison | Aitchison centre  |
|                             | Euclidean     | cosine    | root mean square  |
|                             |               | angular   | root mean square  |
|                             | Chebyshev     | [none]    | max               |
| Full-range (e.g., timbre)   | [none]        | Euclidean | mean              |
|                             | Euclidean     | cosine    | root mean square  |
|                             |               | angular   | root mean square  |


## Bloed, Zweet en Tranen

The following examples from Andre Hazes's ‘Bloed, Zweet en Tranen’ highlight how to use these functions. Can you choose better combinations of normalisations, distances, and summary statistics?

```{r}
bzt <- 
    get_tidy_audio_analysis('5ZLkc5RY1NM4FtGWEd6HOE') %>% 
    compmus_align(bars, segments) %>% 
    select(bars) %>% unnest(bars) %>% 
    mutate(
        pitches = 
            map(segments, 
                compmus_summarise, pitches, 
                method = 'rms', norm = 'euclidean')) %>% 
    mutate(
        timbre = 
            map(segments, 
                compmus_summarise, timbre, 
                method = 'mean'))
```

### Cepstrogram

We can use `compmus_gather_timbre` much like `compmus_gather_chroma` last week to yield a cepstrogram. Try different levels of structure -- Spotify's estimates of beats, bars, or sections -- to see which level is the most meaningful.

```{r}
bzt %>% 
    compmus_gather_timbre %>% 
    ggplot(
        aes(
            x = start + duration / 2, 
            width = duration, 
            y = basis, 
            fill = value)) + 
    geom_tile() +
    labs(x = 'Time (s)', y = NULL, fill = 'Magnitude') +
    scale_fill_viridis_c(option = 'E') +
    theme_classic()
```

### Self-Similarity Matrices

The function `compmus_self_similarity` is a wrapper around `compmus_long_distance` from last week, for the case where the distances are computed form the same track. Try different distance functions to see what is most useful for this track -- and compare your results with a self-similarity matrix based on chroma.

```{r}
bzt %>% 
    compmus_self_similarity(timbre, 'cosine') %>% 
    ggplot(
        aes(
            x = xstart + xduration / 2, 
            width = xduration,
            y = ystart + yduration / 2,
            height = yduration,
            fill = d)) + 
    geom_tile() +
    coord_fixed() +
    scale_fill_viridis_c(option = 'E', guide = 'none') +
    theme_classic() +
    labs(x = '', y = '')
```

