---
title: "Week 11 · Classification"
author: "John Ashley Burgoyne"
date: "13 March 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Set-up

Be sure to update your installation of compmus before running this vignette. 

We will be using the developing `tidymodels` framework for integrating with the different machine-learning libraries in a consistent manner.

```{r}
library(tidyverse)
library(tidymodels)
library(spotifyr)
library(compmus)
```

In order for the code below to run, it is also necessary to set up Spotify login credentials for `spotifyr`.

## Novelty Functions

For novelty functions, we want to work directly with the segments, and not summarise them at higher levels like Spotify's own estimates of bar or beat.

```{r}
pata_pata <- 
    get_tidy_audio_analysis('3uy90vHHATPjtdilshDQDt') %>% 
    select(segments) %>% unnest(segments)
```

We can compute an energy-based novelty function based on Spotify's loudness estimates. The tempo of this piece is about 126 BPM: how well does this technique work?

```{r}
pata_pata %>% 
    mutate(loudness_max_time = start + loudness_max_time) %>% 
    arrange(loudness_max_time) %>% 
    mutate(delta_loudness = loudness_max - lag(loudness_max)) %>% 
    ggplot(aes(x = loudness_max_time, y = pmax(0, delta_loudness))) +
    geom_line() +
    xlim(0, 30) +
    theme_minimal() +
    labs(x = 'Time (s)', y = 'Novelty')
```

We can use similar approaches for chromagrams and cepstrograms. In the case of chromagrams, Aitchison's clr transformation gives more sensible differences between time points. Even with these helpful transformations, however, self-similarity matrices tend to be more helpful visualisations of chroma and timbre from the Spotify API.

```{r}
pata_pata %>% 
    mutate(pitches = map(pitches, compmus_normalise, 'clr')) %>% 
    arrange(start) %>% 
    mutate(pitches = map2(pitches, lag(pitches), `-`)) %>% 
    compmus_gather_chroma %>% 
    ggplot(
        aes(
            x = start + duration / 2, 
            width = duration, 
            y = pitch_class, 
            fill = pmax(0, value))) + 
    geom_tile() +
    scale_fill_viridis_c(option = 'E', guide = 'none') +
    xlim(0, 30) +
    labs(x = 'Time (s)', y = NULL, fill = 'Magnitude') +
    theme_classic()
```


```{r}
pata_pata %>% 
    arrange(start) %>% 
    mutate(timbre = map2(timbre, lag(timbre), `-`)) %>% 
    compmus_gather_timbre %>% 
    ggplot(
        aes(
            x = start + duration / 2, 
            width = duration, 
            y = basis, 
            fill = pmax(0, value))) + 
    geom_tile() +
    scale_fill_viridis_c(option = 'E', guide = 'none') +
    xlim(0, 30) +
    labs(x = 'Time (s)', y = NULL, fill = 'Magnitude') +
    theme_classic()
```

Find a Spotify track that has a regular tempo but lacks percussion (e.g., much Western classical music), and compute the above three representations. How do they differ from what you see for 'Pata Pata'?

## Tempograms

Spotify does not make the novelty function underlying their own tempo analysis available to the public, but we can still use onsets of every segment to generate Fourier tempograms. The `tempogram()` function from `compmus` generates this automatically from an audio analysis, ready to plot with `geom_raster` (a faster version of `geom_tile` for when every segment has the same length). Here is an example of 'Samba do outro lugar', a track from the Brazilian indie band Graveola that features several tempo and metre alternations.

```{r}
graveola <- get_tidy_audio_analysis('6PJasPKAzNLSOzxeAH33j2')
```

```{r}
graveola %>% 
    tempogram(window_size = 8, hop_size = 1, cyclic = FALSE) %>% 
    ggplot(aes(x = time, y = bpm, fill = power)) + 
    geom_raster() + 
    scale_fill_viridis_c(guide = 'none') +
    labs(x = 'Time (s)', y = 'Tempo (BPM)') +
    theme_classic()
```

The textbook notes that Fourier-based tempograms tend to pick up strongly on tempo harmonics. Wrapping into a cyclic tempogram can be more informative. 

```{r}
graveola %>% 
    tempogram(window_size = 8, hop_size = 1, cyclic = TRUE) %>% 
    ggplot(aes(x = time, y = bpm, fill = power)) + 
    geom_raster() + 
    scale_fill_viridis_c(guide = 'none') +
    labs(x = 'Time (s)', y = 'Tempo (BPM)') +
    theme_classic()
```


## Classification

In order to demonstrate some of the principles of classification, we will try to identify some of the features that Spotify uses to designate playlists as 'workout' playlists. For a full analysis, we would need to delve deeper, but let's start with a comparison of three playlists: Indie Pop, Indie Party, and Indie Workout. For speed, this example will work with only the first 20 songs from each playlist, but you should feel free to use more if your computer can handle it.

```{r}
pop <- 
    get_playlist_audio_features('spotify', '37i9dQZF1DWWEcRhUVtL8n') %>% 
    slice(1:20) %>% 
    add_audio_analysis
party <- 
    get_playlist_audio_features('spotify', '37i9dQZF1DWTujiC7wfofZ') %>% 
    slice(1:20) %>% 
    add_audio_analysis
workout <- 
    get_playlist_audio_features('spotify', '37i9dQZF1DXaRL7xbcDl7X') %>% 
    slice(1:20) %>% 
    add_audio_analysis
```

As you think about this lab session -- and your portfolio -- think about the four kinds of validity that Sturm and Wiggins discussed in our reading for this week. Do these projects have:

  - Statistical validity [somewhat beyond the scope of this course]?
  - Content validity?
  - Internal validity?
  - External validity?

We bind the three playlists together using the trick from Week 7, transpose the chroma vectors to a common tonic using the `compmus_c_transpose` function, and then summarise the vectors like we did when generating chromagrams and cepstrograms. Again, Aitchison's clr transformation can help with chroma.

```{r}
indie <- 
    pop %>% mutate(playlist = "Indie Pop") %>% 
    bind_rows(
        party %>% mutate(playlist = "Indie Party"),
        workout %>% mutate(playlist = "Indie Workout")) %>% 
    mutate(playlist = factor(playlist)) %>% 
    mutate(
        segments = 
            map2(segments, key, compmus_c_transpose)) %>% 
    mutate(
        pitches = 
            map(segments, 
                compmus_summarise, pitches, 
                method = 'mean', norm = 'manhattan'),
        timbre =
            map(
                segments,
                compmus_summarise, timbre,
                method = 'mean')) %>% 
    mutate(pitches = map(pitches, compmus_normalise, 'clr')) %>% 
    mutate_at(vars(pitches, timbre), map, bind_rows) %>% 
    unnest(cols = c(pitches, timbre))
```

### Pre-processing

In the `tidyverse` approach, we can preprocess data with a `recipe` specifying what we are predicting and what variables we think might be useful for that prediction. Then we use `step` functions to do any data clean (usually centering and scaling, but `step_range` is a viable alternative that squeezes everything to be between 0 and 1). Finally we `prep` and `juice` the data. 

```{r}
indie_class <- 
    recipe(playlist ~
               danceability +
               energy +
               loudness +
               speechiness +
               acousticness +
               instrumentalness +
               liveness +
               valence +
               tempo +
               track.duration_ms +
               C + `C#|Db` + D + `D#|Eb` +
               E + `F` + `F#|Gb` + G +
               `G#|Ab` + A + `A#|Bb` + B +
               c01 + c02 + c03 + c04 + c05 + c06 +
               c07 + c08 + c09 + c10 + c11 + c12,
           data = indie) %>% 
    step_center(all_predictors()) %>%
    step_scale(all_predictors()) %>%
    # step_range(all_predictors()) %>% 
    prep(indie) %>% 
    juice
```

### Cross-Validation

The `vfold_cv` function sets up cross-validation. We will use 5-fold cross-validation here in the interest of spped, but 10-fold cross-validation is more typical. 

```{r}
indie_cv <- indie_class %>% vfold_cv(5)
```

### Classification Algorithms

Your DataCamp tutorials this week introduced four classical algorithms for classification: $k$-nearest neighbour, naive Bayes, logistic regression, and decision trees. Other than naive Bayes, all of them can be implemented more simply in `tidymodels`. In order to use cross-validation, however, we need to write some local helper functions to `fit` the classifier on the training sets, `predict` the labels for the test/validation sets, and `bind` the results to the original data.

#### $k$-Nearest Neighbour

A $k$-nearest neighbour classifier often works just fine with only one neighbour. It is very sensitive to the choice of features, however. Let's check the performance as a baseline and come back to it later.

```{r}
indie_knn <- 
  nearest_neighbor(mode = 'classification', neighbors = 1) %>% 
  set_engine('kknn')
predict_knn <- function(split)
    fit(indie_knn, playlist ~ ., data = analysis(split)) %>% 
    predict(assessment(split), type = 'class') %>%
    bind_cols(assessment(split))
```

After a little awkwardness with cross-validation, we can use `conf_mat` to get a confusion matrix.

```{r}
indie_cv %>% 
    mutate(pred = map(splits, predict_knn)) %>% unnest(pred) %>% 
    conf_mat(truth = playlist, estimate = .pred_class)
```

These matrices `autoplot` in two forms.

```{r}
indie_cv %>% 
    mutate(pred = map(splits, predict_knn)) %>% unnest(pred) %>% 
    conf_mat(truth = playlist, estimate = .pred_class) %>% 
    autoplot(type = 'mosaic')
```

```{r}
indie_cv %>% 
    mutate(pred = map(splits, predict_knn)) %>% unnest(pred) %>% 
    conf_mat(truth = playlist, estimate = .pred_class) %>% 
    autoplot(type = 'heatmap')
```

We can also compute statistics like accuracy, Cohen's kappa, or the J-measure.

```{r}
indie_cv %>% 
    mutate(pred = map(splits, predict_knn)) %>% unnest(pred) %>% 
    metric_set(accuracy, kap, j_index)(truth = playlist, estimate = .pred_class)
```

#### Logistic and Multinomial Regression

In the two-class case, we use logistic regression, but beware if you have more than two classes! R will just build a classifier for the first two without warning.

```{r}
indie_logistic <- logistic_reg() %>% set_engine('glm')
predict_logistic <- function(split)
    fit(indie_logistic, playlist ~ ., data = analysis(split)) %>% 
    predict(assessment(split), type = 'class') %>%
    bind_cols(assessment(split))
```

With three or more classes, we need multinomial regression instead. You can adjust the penalty parameter if you are feeling adventurous.

```{r}
indie_multinom <- multinom_reg(penalty = 0.1) %>% set_engine('glmnet')
predict_multinom <- function(split)
    fit(indie_multinom, playlist ~ ., data = analysis(split)) %>% 
    predict(assessment(split), type = 'class') %>%
    bind_cols(assessment(split))
```

It is not a strong classifier for this problem.

```{r}
indie_cv %>% 
    mutate(pred = map(splits, predict_multinom)) %>% unnest(pred) %>% 
    metric_set(accuracy, kap, j_index)(truth = playlist, estimate = .pred_class)
```

We can look at the most important features in the model by using the `coef` method.

```{r}
indie_class %>% 
    fit(indie_multinom, playlist ~ ., data = .) %>% 
    pluck('fit') %>%
    coef
```

#### Decision Trees

Decision trees are nicely intuitive, and perform somewhat better here.

```{r}
indie_tree <- decision_tree(mode = 'classification') %>% set_engine('C5.0')
predict_tree <- function(split)
    fit(indie_tree, playlist ~ ., data = analysis(split)) %>% 
    predict(assessment(split), type = 'class') %>%
    bind_cols(assessment(split))
```

```{r}
indie_cv %>% 
    mutate(pred = map(splits, predict_tree)) %>% unnest(pred) %>% 
    metric_set(accuracy, kap, j_index)(truth = playlist, estimate = .pred_class)
```

We can look at the whole tree with the `summary` command. Be careful not to read too much into the actual numerical values, however: remember that the features were standardised before we started classification. Without cross-validation, the algorithm looks much better from the summary than it actually was in practice, but we can still see that timbre features are important and chroma features probably aren't. 

```{r}
indie_class %>% 
    fit(indie_tree, playlist ~ ., data = .) %>% 
    pluck('fit') %>%
    summary
```

#### Random Forests

```{r}
indie_forest <- 
  rand_forest(mode = 'classification') %>% set_engine('randomForest')
predict_forest <- function(split)
    fit(indie_forest, playlist ~ ., data = analysis(split)) %>% 
    predict(assessment(split), type = 'class') %>%
    bind_cols(assessment(split))
```

```{r}
indie_cv %>% 
    mutate(pred = map(splits, predict_forest)) %>% 
    unnest(pred) %>% 
    metric_set(accuracy, kap, j_index)(truth = playlist, estimate = .pred_class)
```

Random forests give us the best-quality ranking of feature importance, and we can plot it with `randomForest::varImpPlot`. Again, it is clear that timbre, specifically Component 1 (power) and Component 11, is important. Note that because random forests are indeed random, the accuracy and feature rankings will vary (slightly) every time you re-run the code.

```{r}
indie_class %>% 
    fit(indie_forest, playlist ~ ., data = .) %>% 
    pluck('fit') %>% 
    randomForest::varImpPlot()
```

#### Feature Selection

Let's try $k$-NN again with just the top features. We see much better results.

```{r}
predict_knn_reduced <- function(split)
    fit(
        indie_knn, 
        playlist ~ c01 + c11 + liveness + energy + acousticness, 
        data = analysis(split)) %>% 
    predict(assessment(split), type = 'class') %>%
    bind_cols(assessment(split))
indie_cv %>% 
    mutate(pred = map(splits, predict_knn_reduced)) %>% unnest(pred) %>% 
    metric_set(accuracy, kap, j_index)(truth = playlist, estimate = .pred_class)
```

```{r}
indie_cv %>% 
    mutate(pred = map(splits, predict_knn_reduced)) %>% unnest(pred) %>% 
    conf_mat(truth = playlist, estimate = .pred_class) %>% 
    autoplot(type = 'mosaic')
```

Armed with this feature set, perhaps we can make a better plot. It's clear that the workout list has fewer live tracks, and that the party playlist is somewhat louder and higher on Component 11 than the pop list.

```{r}
indie %>%
    ggplot(aes(x = c01, y = c11, colour = playlist, size = liveness)) +
    geom_point(alpha = 0.8) +
    scale_color_brewer(type = 'qual', palette = 'Accent') +
    labs(x = 'Timbre Component 1', y = 'Timbre Component 11', size = 'Liveness', colour = 'Playlist')
```

## Deltas and Delta-Deltas (Advanced)

Although the novelty-based transformations of chroma and timbre features are not always useful for visualisations, they can be very useful for classification. Both 'deltas' and 'delta-deltas', especially for timbre features, are in regular use in music information retrieval. The code example below shows how to compute average *delta* chroma and timbre features instead of the ordinary average. Can you incorporate it into the classifiers above? Can you add delta-deltas, too?

```{r}
indie_deltas <-
    pop %>% mutate(playlist = "Indie Pop") %>% 
    bind_rows(
        party %>% mutate(playlist = "Indie Party"),
        workout %>% mutate(playlist = "Indie Workout")) %>% 
    mutate(playlist = factor(playlist)) %>% 
    mutate(
        segments = 
            map2(segments, key, compmus_c_transpose)) %>% 
    mutate(
        segments = 
            map(
                segments, 
                mutate, 
                pitches = map(pitches, compmus_normalise, 'manhattan'))) %>% 
    mutate(
        segments =
            map(
                segments,
                mutate,
                pitches = map2(pitches, lag(pitches), `-`))) %>% 
    mutate(
        segments =
            map(
                segments,
                mutate,
                timbre = map2(timbre, lag(timbre), `-`))) %>% 
    mutate(
        pitches =
            map(segments,
                compmus_summarise, pitches,
                method = 'mean', na.rm = TRUE),
        timbre =
            map(
                segments,
                compmus_summarise, timbre,
                method = 'mean', na.rm = TRUE)) %>%
    mutate_at(vars(pitches, timbre), map, bind_rows) %>% 
    unnest(cols = c(pitches, timbre))
```
