---
title: "Week 12 · Similarity and Clustering"
author: "John Ashley Burgoyne"
date: "20 March 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Week 12 · Similarity and Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Set-up

All of the tools that are strictly necessary for clustering are available in base R. For full flexibility, however, the `protoclust` packages is recommended. If you want to explore further possibilities, look at the `cluster` package.

```{r}
library(tidyverse)
library(tidymodels)
library(protoclust)
library(spotifyr)
library(compmus)
```

In order for the code below to run, it is also necessary to set up Spotify login credentials for `spotifyr`.

## Clustering

The Bibliothèque nationale de France (BnF) makes a large portion of its [music collection](https://gallica.bnf.fr/html/und/bnf-collection-sonore) available on Spotify, including an eclectic collection of curated playlists. The defining musical characteristics of these playlists are sometimes unclear: for example, they have a Halloween playlist. Perhaps clustering can help us organise and describe what kinds of musical selections make it into the BnF's playlist.

We begin by loading the playlist and summarising the pitch and timbre features, just like last week. Note that, also like last week, we use `compmus_c_transpose` to transpose the chroma features so that -- depending on the accuracy of Spotify's key estimation -- we can interpret them as if every piece were in C major or C minor. Although this example includes no delta features, try adding them yourself if you are feeling comfortable with R!

```{r}
halloween <- 
    get_playlist_audio_features('bnfcollection', '1vsoLSK3ArkpaIHmUaF02C') %>% 
    add_audio_analysis %>% 
    mutate(
        segments = 
            map2(segments, key, compmus_c_transpose)) %>% 
    mutate(
        pitches = 
            map(segments, 
                compmus_summarise, pitches, 
                method = 'mean', norm = 'manhattan'),
        timbre =
            map(
                segments,
                compmus_summarise, timbre,
                method = 'mean')) %>% 
    mutate(pitches = map(pitches, compmus_normalise, 'clr')) %>% 
    mutate_at(vars(pitches, timbre), map, bind_rows) %>% 
    unnest(cols = c(pitches, timbre))
```

### Pre-processing

Remember that in the `tidyverse` approach, we can preprocess data with a `recipe`. In this case, instead of a label that we want to predict, we start with a label that will make the cluster plots readable. For most projects, the track name will be the best choice (although feel free to experiment with others). The code below uses `str_trunc` to clip the track name to a maximum of 20 characters, again in order to improve readability. The other change from last week is `column_to_rownames`, which is necessary for the plot labels to appear correctly.

Last week we also discussed that although standardising variables with `step_center` to make the mean 0 and `step_scale` to make the standard deviation 1 is the most common approach, sometimes `step_range` is a better alternative, which squashes or stretches every features so that it ranges from 0 to 1. For most classification algorithms, the difference is small; for clustering, the differences can be more noticable. It's wise to try both.

```{r}
halloween_juice <- 
    recipe(track.name ~
               danceability +
               energy +
               loudness +
               speechiness +
               acousticness +
               instrumentalness +
               liveness +
               valence +
               tempo +
               track.duration_ms +
               C + `C#|Db` + D + `D#|Eb` +
               E + `F` + `F#|Gb` + G +
               `G#|Ab` + A + `A#|Bb` + B +
               c01 + c02 + c03 + c04 + c05 + c06 +
               c07 + c08 + c09 + c10 + c11 + c12,
           data = halloween) %>% 
    step_center(all_predictors()) %>%
    step_scale(all_predictors()) %>%
    # step_range(all_predictors()) %>% 
    prep(halloween %>% mutate(track_name = str_trunc(track.name, 20))) %>% 
    juice %>% 
    column_to_rownames('track.name')
```

### Computing distances

When using `step_center` and `step_scale`, then the Euclidean distance is usual. When using `step_range`, then the Manhattan distance is also a good choice: this combination is known as *Gower's distance* and has a long history in clustering.

```{r}
halloween_dist <- dist(halloween_juice, method = 'euclidean')
```

### Hierarchical clustering

As you learned in your DataCamp exercises this week, there are three primary types of *linkage*: single, average, and complete. Usually average or complete give the best results. We can use the `ggendrogram` function to make a more standardised plot of the results.

```{r}
hclust(halloween_dist, method = 'single') %>% plot
```

A more recent -- and often superior -- linkage function is *minimax linkage*, available in the `protoclust` package. It is more akin to $k$-means: at each step, it chooses an ideal centroid for every cluster such that the maximum distance between centroids and all members of their respective clusters is as small as possible.

```{r}
protoclust(halloween_dist) %>% plot
```

Try all four of these linkages. Which one looks the best? Which one *sounds* the best (when you listen to the tracks on Spotify)? Can you guess which features are separating the clusters? 

### *k*-Means

Unlike hierarchical clustering, *k*-means clustering returns a different results every time. Nonetheless, it can be a useful reality check on the stability of the clusters from hierarchical clustering.

```{r}
kmeans(halloween_juice, 4)
```